---
title: "R Guide to ANOVA"
author: "Joe Hilgard"
date: "December 28, 2017"
output: html_document
---

https://stackoverflow.com/questions/45140783/how-to-transform-a-string-into-a-factor-and-sets-contrasts-using-dplyr-magrittr

https://pagepiccinini.com/2016/03/18/contrast-coding-with-three-level-variables/

Working with factors in R can be unpleasant. Additionally, it is not always clear what the output means when you run summary() or car::Anova() on an lm() or aov() object.

This guide is meant to clarify a few sticking points in working with factors. We will use the mtcars dataset as an example throughout.

mtcars has two 0, 1 columns, vs and am. We will use these as factors so that we can turn it into a 2×2 ANOVA example. 

To make things clearer, we will convert them into text labels. The first text label becomes the label of the first factor level, and the second text label becomes the label of the second factor level.

```{r}
library(tidyverse)
# Make default factor objects
dat <- mtcars %>% 
  mutate(vs = factor(vs, labels = c("v", "s")),
         am = factor(am, labels = c("automatic", "manual")),
         cyl = factor(cyl, labels = c("four", "six", "eight")))
# Make factor objects with contr.sum
dat <- dat %>% 
  mutate(vs.c = vs,
         am.c = am,
         cyl.c = cyl)
contrasts(dat$vs.c)<- contr.sum
contrasts(dat$am.c)<- contr.sum
contrasts(dat$cyl.c)<- contr.sum
```
We can check that this went as planned by comparing the original mtcars columns against the new ones.

```{r}
data.frame(mtcars$vs, dat$vs, mtcars$am, dat$am) %>% 
  head()
```

Okay. Let's just charge on ahead and run our 2x2 ANOVA.

```{r}
m1 <- lm(mpg ~ am * vs, data = dat)
summary(m1)
```
So in the dummy code analysis, the intercept is 15. This is the cell mean of the am = 0, vs = 0 group, which in our case is automatic cars with a v engine. The am coefficient tells us that relative to our v-engine automatics, v-engine manuals have 4.7 mpg more, on average. (Notice that we are still talking about v engines -- simple effects!) The vs coefficient indicates that a straight engine automatic gets 5.7 more miles per gallon. The interaction term gets us to the last cell mean: manual transmissions with a straight engine.

It's important to know that exactly these effects represent. Many users, in my experience, expect the default parameterization to resemble that of ANOVA

```{r}
m2 <- aov(mpg ~ am * vs, data = dat)
summary(m2)
```

If you want the lm-style coefficients, we can get that same output by running lm() but using contrast codes:

```{r}
m3 <- lm(mpg ~ am.c * vs.c, data = dat)
summary(m3)
```

Look at this -- now the p-values are a match!

Now the intercept represents the grand mean of 21 mpg across all vehicles. An automatic subtracts 3, and a manual adds 3. A v-engine subtracts 3.6, and a straight adds 3.6. And if you have a manual straight-engine, you add 3, add 3.6, then subtract 0.7, the interaction term.

"Uh oh!" you say, "aov() and lm() give different p-values! Which is right?" The reason for this is that these coefficients represent different things. aov() yields the main effects of am and vs, but lm(), by default, yields the simple effects.

What's the difference between main and simple effects? A *main effect* is the effect marginalized across all other effects. When we talk about the main effect of a manual transmission (as compared to an automatic transmission), we are implicitly averaging across v-engines and straight-engines. 

A *simple effect* is the effect at a *specific level of the other factors*. Because there are two levels of vs, there are two simple effects of transmission: one simple effect given that the car has a v-engine and another simple effect given that the car has a straight engine.

Why does aov() give the main effect and lm() give the simple effect? And can lm() give the main effect? The diference between these two functions comes down to the *contrasts* --- how R parameterizes the model. By default, lm() uses  what it calls "treatment contrasts". These treatment contrasts labeling data as 0 or 1 depending on whether it is in the group or not. In psychology, we often call these "dummy codes"; in computer science, they have the sexier name "one-hot codes". 

Here's what this looks like for a factor with two levels.
```{r}
contr.treatment(2)
```

Here's what this looks like for a factor with three levels.
```{r}
contr.treatment(3)
```

Four levels, and you get the idea.
```{r}
contr.treatment(4)
```

aov(), on the other hand, uses what it calls "sum contrasts", labeling data as -1, 0, or 1 according to its group. In psychology, we usually call these *contrast codes* or *effect codes*. 

Here's what this looks like for a factor with 2 levels.
```{r}
contr.sum(2)
```

Here's what this looks like for a factor with 3 levels.
```{r}
contr.sum(3)
```

Here's what this looks like for a factor with 4 levels. You get the idea.
```{r}
contr.sum(4)
```


This choice of parameterization has some consequences for what the output means. With dummy codes, the intercept represents the cell mean of some reference group, and the coefficients are each simple effects describing how groups differ from the reference group. With contrast codes, on the other hand, the intercept represents the grand mean of all observations, and the coefficients are main effects describing how groups differ from the grand mean.

With contrast codes, on the other hand, the intercept represents the grand mean. The coefficients, then, represent some displacement from the grand mean.





It's unclear sometimes what to expect from R's parameterization. Is the first element 0 and the second element 1, as would be expected from a zero-based array? Or is the first element 1 and the second element 0, as you would expect if the last level were treated as a reference group? When are attributes assigned by alphabetical order? numerical order? factor level order?

<<<<<<< HEAD

Which parameterization do you want to run, and how can you set that parameterization?

To change an object's contrasts, store the new contrast type to the result of its contrasts() call.
```{r}
contrasts(dat$am)               # look at existing contrast
contrasts(dat$am) <- contr.sum  # assign contr.sum coding instead
contrasts(dat$am)               # look at new contrast codes
```

You can check on what values have which contrast codes by calling contrasts() or just printing the column.
```{r}
# dummy coded version
contrasts(dat$vs)
contrasts(dat$am)
# contrast coded version
contrasts(dat$vs.c)
contrasts(dat$am.c)
```

